Discriminating the Influence of Correlated Factors \\from Multivariate Observations: the Back-to-Back Regression

### 2020-05-19 paper/synthetic.tex
synthetic

We evaluate the performance of B2B throughout a series of experiments on
controlled synthetic data.
%
The purpose of these experiments is to evaluate the ability of B2B on its
ability to 1) accurately predict independent and identically distributed data,
and 2) recover causal factors when the ground truth is known.

@@

\subsection{Baseline models}

All baseline method were based on the implementations of scikit-learn \citep{sklearn} and pyrcca
\citep{bilenko2016pyrcca}. For pedagogical purposes, we briefly summarize them below.

@@

\subsection{Evaluating Causal Discovery from models' coefficients}

%
B2B leads to \emph{scalar} coefficients for non-causal
features. The diagonal of this matrix, $\hat E$, can thus be directly used
as a causal contribution estimate. Note that is estimate is unbiased (i.e.
zeros-centered) if the second regression $H$ is not regularized.

In contrast, the Forward, Backward, CCA and PLS models lead to a
loading \emph{vector} for each feature: each vector indicates how to it
relates to each dimension of the observations (for forward or backward
models) or each canonical projection of the observations
To an estimated causal contribution, we must thus transform such vector into a
scalar, by e.g. taking the sum of square coefficients:
% \begin{equation}
  $\hat E_i = \sum_j {H^j_i}^2 $
% \end{equation}
Note that in such B2B cases, the estimates are necessarily biased: i.e. positive.

Finally, to estimate whether each model accurately identifies causal factors independently of
their potential biases, we compute the
area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
%\begin{equation} AUC = 1 - \sum_1^n (E_k - E_{k-1}) ( \hat{E}_k +
%\hat{E}_{k-1}) / 2 \end{equation}
By definition, this AUC evaluation can only be done when ground truth labels are available, as is the case in
this synthetic setup.

We report AUC results in Figures~\ref{fig:percondition}~(top) and ~\ref{fig:auc_plots}~(left, in Appendix), and compare favorably to all baselines.

@@

\subsection{Evaluating Causal Discovery with held-out prediction reliability}

In most cases, $E$ is not known. The above AUC can thus not be estimated.
To address this issue, we assess the ability of each model to reliably predict
independent and identically distributed data from $Y$, given all of the $X$
features versus all-but-one feature $X_{-i}$ (i.e. 'knock-out X'). This
procedure results in two correlation metrics $R_{full}$ and $R_{knockout}$ for
each feature (for the B2B and Backward models) for each dimension of $Y$ (for
the Forward model) or each canonical dimension of $Y$ (for CCA and PLS). The
difference $\Delta R_i = R_{full}-R_{knockout}$ indicates how much each
$X_i$ improves the prediction of (a pre-defined dimension or projection of) $Y$.

@@

\subsection{Results of the synthetic experiment}

Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in
Appendix) show the results of our synthetic experiment. Note that the figure
does not display each feature $i$ separately, as they are
randomly generated. $\Delta R$ represents the average of $\Delta R_i$. A
higher score means that for prediction, the model relies on individual features
rather than combinations of features. Overall both the AUC and the held-out
prediction reliability evaluations show that B2B outperforms baseline models.

>>>>

We evaluate the performance of B2B throughout a series of experiments on
controlled synthetic data.
%
The purpose of these experiments is to evaluate the ability of B2B on its
ability to 1) recover causal factors when the ground truth is known and 2)
accurately predict independent and identically distributed data otherwise.

@@

\subsubsection{Baseline models}

All baseline methods were based on the implementations of scikit-learn \citep{sklearn} and pyrcca
\citep{bilenko2016pyrcca}. For pedagogical purposes, we briefly summarize them below.

@@

\subsubsection{Evaluating Causal Discovery from models' coefficients}

%
B2B leads to \emph{scalar} coefficients for non-causal
features. The diagonal of this matrix, $\hat E$, can thus be directly used
as a causal contribution estimate. Note that this estimate is unbiased (i.e.
zeros-centered) only if the second regression $H$ is not regularized.

In contrast, the Forward, Backward, CCA and PLS models lead to a
loading \emph{vector} for each feature: each vector indicates how to a feature
relates to each dimension of the observations (for Forward or Backward
models) or each canonical projection of the observations.
To estimate causal contribution, we must thus transform such vector into a
scalar, by e.g. taking the sum of square coefficients:
% \begin{equation}
  $\hat E_i = \sum_j {H^j_i}^2 $
% \end{equation}
Note that in such B2B cases, the estimates are  thus necessarily positive and thus biased.

Finally, to estimate whether each model accurately identifies causal factors independently of
their potential biases, we compute the
area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
%\begin{equation} AUC = 1 - \sum_1^n (E_k - E_{k-1}) ( \hat{E}_k +
%\hat{E}_{k-1}) / 2 \end{equation}
By definition, this AUC evaluation can only be done when ground truth labels are available, as is the case in
this synthetic setup, but not in the neuroimaging experiments below.

@@

\subsubsection{Evaluating Causal Discovery with held-out prediction reliability}

In most real-world cases, $E$ is not known. Consequently, the above AUC evaluation can not be estimated.
To address this issue, we assess the ability of each model to reliably predict
independent and identically distributed data from $Y$, given all of the $X$
features versus all-but-one feature $X_{-i}$ (i.e. 'knock-out X'). This
procedure results in two correlation metrics $R_{full}$ and $R_{knockout}$ for
each feature (for the B2B and Backward models) for each dimension of $Y$ (for
the Forward model) or each canonical dimension of $Y$ (for CCA and PLS). The
difference $\Delta R_i = R_{full}-R_{knockout}$ indicates how much each
$X_i$ improves the prediction of (a pre-defined dimension or projection of) $Y$.

@@

\subsubsection{Results of the synthetic experiment}

Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in
Appendix) show the results of our synthetic experiment. Note that the figure
does not display each feature separately, as they are
randomly generated. $\Delta R$ represents the average of $\Delta R_i$ across
features $i$.
Overall both the AUC and the held-out prediction reliability evaluations
show that B2B compares favorably to the baseline models.
